Episodic memory: mental time travel or a quantum `memory wave' function?

### 2020-09-22 paper/main.tex
minor updates to QWF section
finished draft addressing r1 comments, working on r2 comments [WIP]

\section*{Defining the ``quantum memory wave function'' metaphor and comparing it to the ``mental time travel'' metaphor}
Unlike in classical physics, where matter dutifully obeys Newton's laws, quantum physics postulates a different (though potentially complementary) set of rules governing how matter behaves and interacts.  Perhaps the most famous equation in quantum physics is \cite{Schr26}'s equation describing the quantum \textit{wave function}~\citep[this finding was also inspired by][]{deBr24}.  Schr\"{o}dinger's equation describes, in probabilistic terms, the location of a particle in space, at a given time.  In other words, it defines a function, $\Psi(x, t)$, such that the probability that a particle is located at position $x$ at time $t$ is equal to $|\Psi(x, t)|^2$.  The central idea is that, although it may not be possible to know or predict the particle's precise position at a given moment, the wave function tells us how it's location's probabilty mass extends over space.  In turn, this may allow one to use the wave function to predict behaviors or constraints of the system(s) in which the particle is behaving and interacting.  Within the domain of episodic memory, for the purposes of describing the \textit{quantum memory wave function} metaphor, we can conceptualize the moments along our \textit{autobiographical timeline} as the spatial coordinates ($x$) in Schr\"{o}dinger's equation.  Then our mental state at time $t$ may be conceptualized as spreading its probability mass over our autobiographical timeline.  In other words, like Schr\"{o}dinger's equation, we could in principle define a function describing how our mental state at time $t$ reflects the blends of thoughts we experienced at other moments of our lives.  Under this metaphor, the effect of mental operations we perform with respect to our autobiographical timeline, such as remembering past experiences or predicting what will happen in the future, is to shift the probability mass of our quantum memory wave function (Fig.~\ref{fig:qwave}).

@@

The way they described the example event did \textit{not} mirror the way they described other semantically (but not conceptually) related events, such as events involving unrelated violence, or Sherlock and John interacting with other characters about other subjects.  Taken together, this suggests that these specific events were associated in participants' memories, perhaps contributing to (or reflecting) their understanding that those events were linked in the episode's narrative.

@@

@@

Early word embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{DeerEtal90, LandDuma97} attempted to compute these embeddings using word co-occurance statistics.  The logic driving these approaches posited that documents that contained similar mixtures of words would be likely to be conceptually similar, and words that appeared in similar documents would also likely be conceptually similar.  To quantify these intuitions, \cite{LandDuma97} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of time a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents by number-of-words word count matrix into the product of a number-of-documents $\times K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K \times $ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflected the similarities between documents, and the columns of the word embeddings matrix reflected similarities between words.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.75\textwidth]{figs/word_embedding_models}
\caption{\textbf{Three approaches to text embedding.}  \textbf{A.~Matrix factorization models.} Per-document (or per-participant) word counts and/or frequencies are collected into a word count matrix (left).  To compute $K$-dimensional embeddings of the documents and words, the word count matrix is factorized (typically by computing its Singular Value Decomposition) into the product of a number-of-documents $\times K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K \times $ number-of-unique-words \textit{word embeddings matrix}.  \textbf{B.~Topic models.}  Fitting a topic model to a word count matrix (e.g., Panel A) reveals a set of latent \textit{topics}, or themes that pervade each document.  The topics are specified in a $K \times$ number-of-words \textit{topics matrix}, and the per-document themes are specified in the number-of-documents $\times K$ \textit{topic proportions matrix}. A given document's topic proportions reflect the specific mix of themes that are reflected in that document (bar graph on the left).  These proportions are discovered automatically by estimating the topic from which each word in the document is most probably drawn (highlighted words in \textit{Document text}).  \textbf{C.~Deep learning models.}  A wide variety of deep learning approaches to text embedding have been proposed.  Broadly, these approaches often combine a common set of elements, including \textit{transformers}~\citep[diagram adapted from][]{ViswEtal17}, \textit{deep averaging networks}~\citep[diagram adapted from][]{IyyeEtal15}, \textit{recurrent neural networks}~\citep[diagram adapted from][]{IyyeEtal15}, and \textit{deep autoencoders}~\citep[diagram adapted from][]{YousHame17}.}
\label{fig:embedding-models}
\end{figure}

@@

A key advantage that topic models have over matrix factorization models is the ability to explain potentially ambiguous word usage.  For example, consider the use of the word ``bat'' in the following passage:
\begin{quote}
  \textit{``...As the mutant vampire bat swooped low to attack, the explorer took a mighty swing of her trusty baseball bat.  The swing connected, stunning the creature before it scrambled about on the ground and flew away.  She didn't even bat an eye-- as the dark forces grew in power, such things had become increasingly common, nearly to the point of mundanity...''}
  \end{quote}
At different ponts in the passage, ``bat'' refers to an animal, a piece of sports equipment, and a verb.  Because each of these uses of the word appear in the same document, models driven solely by word co-occurance statistics will blend these different meanings such that the embedding of ``bat'' is somewhat similar to other animals \textit{and} other pieces of sports equipment \textit{and} other eye-related verbs.  By contrast, treating each document as reflecting a blend of topics allows these uses to be disambiguated.  If other documents that mention the word ``bat'' describe vampires, swooping, or other mammals, but \textit{not} sports equipment or eye-related verbs, then a topic model could explain this observation by inferring that multiple topics place substantial weight on the word ``bat.''  Another strength of topic models is their interpretability.  Because the models directly output each document's topic proportions along with each topic's weights on all words in the vocabulary, the general themes that each topic reflects are often intuitive and the per-document topic proportions may be directly related to those themes.

@@

% deep learning approaches:
% - bag of words vs. sequence models vs. sentence-level embeddings
% - accounting for use of context and grammar to convey meaning
% - pre-training vs. customized training vs. training-free

% Neural Probabilistilc Language Model (NPLM): BengEtal03 --> explicitly model word sequences
% Scalable Hierarchical Distributed Language Model (SHDLM): MnihHint09 --> sequences of word trees

%word2vec: MikoEtal13 --> semantic relations reflected in geometric operations
%GloVe: PennEtal14 --> combines matrix factorization and local context window methods
%SkipThought: KiroEtal15 --> reconstructs surrounding sentences of an encoded passage
%BERT: DevlEtal18 --> pretraining language representations, context dependence of word embeddings
%ELMO: PeteEtal18 --> how is it different from BERT/GPT?
%InferSent: ConnEtal18 --> embeddings of sentences
%Universal Sentence Encoder: CerEtal18 --> embeddings of sentences
%Open-GPT: RadfEtal18 --> pretrained transformer (how is it different from BERT?)
%RandSent: WietKiel19 --> training-free sentence-level embeddings
%GPT-2: RadfEtal19
%GPT-3: BrowEtal20

\subsection*{Extensions to spatial, semantic, affective, and social representations}
Just as we mentally spread ourselves across a distribution of times when we remember, we can perform similar mental operations when we conceptualize and retrieve other types of information.  Further, these distributions may be estimated using existing methods.  For example, consider how we use pattern classifiers~\citep{NormEtal06b} to connect neural activity patterns to specific semantic~\citep{PolyEtal05a, MitcEtal08a, MannEtal12}, spatial~\citep{MillEtal13}, affective~\citep{ChanEtal18}, and social~\citep{MeyeEtal18} content being mentally manipulated.
A prevailing approach is to label specific brain activity patterns as each reflecting \textit{one particular} concept or mental state.  Variability in neural activity over repeated experiences or rememberings can result in classifiers spreading their predictions over multiple possible mental states.  Traditional approaches then ``clean up'' this distribution over possible mental states by taking its average (expected), maximum, or modal value.  However, following the ``quantum wave function'' logic described in this manuscript, I suggest that the specific shape of these distributions may not only be informative, but they are \textit{more accurate reflections of what our neural activity patterns actually reflect}.  What are some implications of this view for non-memory domains?

One example of how the full distribution over possible mental states can provide insights into cognition may be seen in how the brain encodes spatial locations.  Theoretical~\citep[e.g., ][]{GersAbbo97} and empirical~\citep[e.g., ][]{PfeiFost13} work has demonstrated how place cell receptive fields may be modulated according to environmental features such as obstacles and the locations of rewarded goals.  In other words, although place cells encode an animal's location within an environment, even for a fixed cell and environment there is some variability with respect to where precisely within that environment the cell modulates its firing rate.  These studies suggest that variability in the set of locations to which a place cell responds is not merely a reflection of noise or imprecision.  Rather, this variability might have some functional relevance in terms of facilitating the learning of paths around obstacles and towards goals.

>>>>

\section*{Defining the ``quantum memory wave function'' metaphor and comparing it to the ``mental time travel'' metaphor}
Unlike in classical physics, where matter dutifully obeys Newton's laws, quantum physics postulates a different (though potentially complementary) set of rules governing how matter behaves and interacts.  Perhaps the most famous equation in quantum physics is \cite{Schr26}'s equation describing the quantum \textit{wave function}~\citep[this finding was also inspired by][]{deBr24}.  Schr\"{o}dinger's equation describes, in probabilistic terms, the location of a particle in space, at a given time.  In other words, it defines a function, $\Psi(x, t)$, such that the probability that a particle is located at position $x$ at time $t$ is equal to $|\Psi(x, t)|^2$.  The central idea is that, although it may not be possible to know or predict the particle's precise position at a given moment, the wave function tells us how it's location's probabilty mass extends over space.  In turn, this may allow one to use the wave function to predict behaviors or constraints of the system(s) in which the particle is behaving and interacting.  Within the domain of episodic memory, for the purposes of describing the \textit{quantum memory wave function} metaphor, we can conceptualize the moments along our \textit{autobiographical timeline} as the spatial coordinates ($x$) in Schr\"{o}dinger's equation.  Then our mental state at time $t$ may be conceptualized as spreading its probability mass over our autobiographical timeline.  In other words, like Schr\"{o}dinger's equation, we could in principle define a function describing how our mental state at time $t$ reflects the blends of thoughts we experienced at other moments of our lives.  Under this metaphor, the effect of mental operations we perform with respect to our autobiographical timeline, such as remembering past experiences or predicting what will happen in the future, is to shift the probability mass of our quantum memory wave function (Fig.~\ref{fig:qwave}).  This metaphor is also related to work on quantum probability decision models~\citep[e.g.,][]{KhreEtal18} in that it extends similar concepts to the domain of episodic memory.

@@

The way they described the example target event did \textit{not} mirror the way they described other events that were not directly narratively linked with the target event, even when those other events overlapped semantically.  For example, thematic elements of events involving unrelated violence, or Sherlock and John interacting with other characters about other subjects, tended not to feature in participants' recountings of the target event.  Taken together, this suggests that these specific events were associated in participants' memories, perhaps contributing to (or reflecting) their understanding that those events were linked in the episode's narrative.

@@

In their paper, \cite{HeusEtal18c} use the geometric shapes of experience trajectories (through word embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal18c} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  This may be characterized by examining the low spatial frequency details of the trajectory's shape.  Low-level details are reflected by (typically smaller-scale) within-event geometric patterns, which may be characterized by examining the high spatial frequency details of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experience relative to others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

@@

Early word embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{DeerEtal90, LandDuma97} attempted to compute these embeddings using word co-occurance statistics.  The logic driving these approaches posited that documents that contained similar mixtures of words would be likely to be conceptually similar, and words that appeared in similar documents would also likely be conceptually similar.  To quantify these intuitions, \cite{LandDuma97} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of time a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents by number-of-words word count matrix into the product of a number-of-documents $\times~K$ \textit{document embeddings matrix}, a $K~\times~K$ \textit{singular values} matrix, and a $K~\times$ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflected the similarities between documents, and the columns of the word embeddings matrix reflected similarities between words.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/word_embedding_models}
\caption{\textbf{Three approaches to text embedding.}  \textbf{A.~Matrix factorization models.} Per-document (or per-participant) word counts and/or frequencies are collected into a word count matrix (left).  To compute $K$-dimensional embeddings of the documents and words, the word count matrix is factorized (typically by computing its Singular Value Decomposition) into the product of a number-of-documents $\times~K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K~\times $ number-of-unique-words \textit{word embeddings matrix}.  \textbf{B.~Topic models.}  Fitting a topic model to a word count matrix (e.g., Panel A) reveals a set of latent \textit{topics}, or themes that pervade each document.  The topics are specified in a $K~\times$ number-of-words \textit{topics matrix}, and the per-document themes are specified in the number-of-documents $\times~K$ \textit{topic proportions matrix}. A given document's topic proportions reflect the specific mix of themes that are reflected in that document (bar graph on the left).  These proportions are discovered automatically by estimating the topic from which each word in the document is most probably drawn (highlighted words in \textit{Document text}).  \textbf{C.~Deep learning models.}  A wide variety of deep learning approaches to text embedding have been proposed.  Broadly, these approaches often combine a common set of elements, including \textit{transformers}~\citep[diagram adapted from][]{ViswEtal17}, \textit{deep averaging networks}~\citep[diagram adapted from][]{IyyeEtal15}, \textit{recurrent neural networks}~\citep[diagram adapted from][]{IyyeEtal15}, and \textit{deep autoencoders}~\citep[diagram adapted from][]{YousHame17}.}
\label{fig:embedding-models}
\end{figure}

@@

A key advantage that topic models have over matrix factorization models is the ability to explain potentially ambiguous word usage.  For example, consider the use of the word ``bat'' in the following illustrative passage:
\begin{quote}
  \textit{``As the slavering mutant vampire bat swooped low to attack, its eyes ablaze, the explorer grunted and took a mighty swing of her trusty aluminum baseball bat.  The powerful swing connected with a echoing metallic \textit{clink}, stunning the creature before it scrambled about on the cave floor and flew away.  She didn't even bat an eye-- as the Dark Wizard Chiroptera continued to draw strength from the mystic crystals, such things had become increasingly common of late, nearly to the point of mundanity...''}
  \end{quote}
At different ponts in the passage, ``bat'' refers to an animal, a piece of sports equipment, and a verb.  Because each of these uses of the word appear in the same document, models driven solely by word co-occurance statistics will blend these different meanings such that the embedding of ``bat'' is somewhat similar to other animals \textit{and} other pieces of sports equipment \textit{and} other eye-related verbs.  By contrast, treating each document as reflecting a blend of topics allows these uses to be disambiguated.  If other documents that mention the word ``bat'' describe vampires, swooping, or other mammals, but \textit{not} sports equipment or eye-related verbs, then a topic model could explain this observation by inferring that multiple topics place substantial weight on the word ``bat.''  Another strength of topic models is their interpretability.  Because the models directly output each document's topic proportions along with each topic's weights on all words in the vocabulary, the general themes that each topic reflects are often intuitive and the per-document topic proportions may be directly related to those themes.

@@

\subsection*{Extensions to spatial, semantic, affective, and social representations}
The way we navigate through \textit{thought space} (i.e., word embedding spaces; Figs.~\ref{fig:trajectories}, \ref{fig:embedding-models}) differs from how we navigate in physical spatial environments.  First, we cannot teleport in real space, whereas our thoughts can exhibit rapid jumps between different non-contiguous regions of word embedding space.  Second, when two people visit the same fixed physical landmark, they necessarily visit the same physical location.  When two people describe a shared experience, their idiosyncratic thoughts, goals, prior experiences, etc. can lead them to process and remember the experience differently.  Third, when we revisit a fixed physical landmark, we (by definition) revisit the same spatial location.  In other words, it is possible to visit the same spatial location twice.  However, we cannot revisit our prior experiences in this way.  We only get to experience each moment once; subsequent ``visits'' to a prior experience must occur through a different medium (e.g., our memory, another person’s memory, a physical recording, a written account, etc.).  Fourth, we can revisit different aspects of the same prior experience on different ``visits''-- e.g., when we remember a specific event we can focus on the physical occurrence in one remembering, the emotional occurence in another remembering, the social implications in another remembering, etc.  We can also revisit the same experience in different levels of detail or depth.  There are no analogs of these phenomena in spatial navigation.  Finally, \textit{physically} (at a macro scale), we are always located at a single moment and in a single location.  The main argument of the quantum memory wave function framework I propose here is that we may be \textit{mentally} spread over many times or locations simultaneously.  For example, this provides a way for us to integrate information from multiple prior experiences into a single decision or conceptualization, even if those experiences were not temporally contiguous.

Many of these differences between physical and mental navigation also apply to other mental domains, including how we \textit{conceptualize} space (e.g., when we trace out a potential route in our mind).  Just as we mentally spread ourselves across a distribution of times when we remember, we can perform similar mental operations when we conceptualize and retrieve other types of information.  Further, these distributions may be estimated using existing methods.  For example, consider how we use pattern classifiers~\citep{NormEtal06b} to connect neural activity patterns to specific semantic~\citep{PolyEtal05a, MitcEtal08a, MannEtal12}, spatial~\citep{MillEtal13}, affective~\citep{ChanEtal18}, and social~\citep{MeyeEtal18} content being mentally manipulated.  A prevailing approach is to label specific brain activity patterns as each reflecting \textit{one particular} concept or mental state.  Variability in neural activity over repeated experiences or rememberings can result in classifiers spreading their predictions over multiple possible mental states, including non-contiguous mental states.  Traditional approaches then ``clean up'' this distribution over possible mental states by taking its average (expected), maximum, or modal value.  However, following the ``quantum wave function'' logic described in this manuscript, these multi-peaked receptive fields may provide a mechanism for associating non-contiguous representations.  What are some implications of this view for non-memory domains?

One example of how the full distribution over possible mental states can provide insights into cognition may be seen in how the brain encodes spatial locations.  Theoretical~\citep[e.g., ][]{GersAbbo97} and empirical~\citep[e.g., ][]{PfeiFost13} work has demonstrated how place cell receptive fields may be modulated according to environmental features such as obstacles and the locations of rewarded goals.  In other words, although place cells encode an animal's location within an environment, even for a fixed cell and environment there is some variability with respect to where precisely within that environment the cell modulates its firing rate.  Further, a number of studies have suggested that some neurons respond to distinct (non-contiguous) places~\citep{FentEtal08, RichEtal14b, LeeEtal19, DerdEtal09, GrieEtal20} and times~\citep{PastEtal08}.  Taken together, these studies suggest that variability in the set of locations and/or times to which a neuron responds is not merely a reflection of noise or imprecision.  Rather, this variability might have some functional relevance in terms of facilitating the learning of paths around obstacles and towards goals.
